A Note on Convergence Diagnostics for Multiple Imputation Algorithms

### 2019-12-14 Simulation/Writeup/NoteConvergence.Rnw
additions to discussion (future research topics)

The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets. This note provides the theoretical foundation towards the diagnostic evaluation of the convergence of MI algorithms. % for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm. 
For reasons of brevity, this note only focuses on the MI algorithm implemented in \pkg{mice} \cite{mice}. %The implementation limits itself to the MI algorithm 
%More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}. 
To this end, we will evaluate how convergence of the imputation algorithm can be diagnosed.

@@

The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problems. Basic familiarity with MI methodology is assumed.\footnote{For the theoretical foundation of MI, see \cite{rubin87}. For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.}  
This note follows notation and conventions of \cite{mice}. % to suit the \pkg{mice} environment. %Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.

@@

To define $\widehat{R}$, we follow notation by \citep[p.~5]{veht19}. % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}
Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance). For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. There is, however, no baseline measure available to evaluate their performance against.
\footnote{The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.} 
We will assess the appropriateness of the two convergence diagnostics in comparison to evaluation criteria recommended by \citep[\S~2.5.2]{buur18}. The performance diagnostics comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.). For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

\subsection{Data simulation}
The point of origin for all simulation conditions is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variable $X$, and covariates $Z_1$ and $Z_2$: $Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2$. The quantity of scientific interest is regression coefficient $\beta_1$. The data generating model is a multivariate normal distribution with means structure $\mu$ and variance-covariance matrix $\Sigma$.

@@

Outcome variable $Y$ is subsequently calculated as $Y =  2X + .5Z_1 - Z_2 + \epsilon$.

@@

\subsection{Convergence Diagnostics}
% Add ``post-processing steps (e.g. aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results''. %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
We investigate convergence by extracting not only the imputed values of each MI procedure (the $T^{th}$ iteration), but also information ($\theta$-values) from the intermediate iterations ($t = 1, 2, \dots, T$). The $\theta$s of interest are chain means and chain variances per variable. We compute convergence diagnostics separately for each of the four variables in the dataset, and report the maximum (absolute) value as the `worst linear predictor' of convergence \textbf{[Rephrase or footnote??]}. $\widehat{R}$ is computed by implementing the definition given in \textbf{EQUATION NR}. $AC$ is computed with \pkg{stats} function \fct{acf}.

@@

We compute bias as the difference between $\bar{Q}$ and $Q$. Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$ (CI95\%). We compute the CI95\% bounds as $\bar{Q} \pm 2.66 \times SE_{\bar{Q}}$, where 2.66 is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.

@@

\textbf{Empirical coverage rate seems more or less stable for conditions where $T>1$. We see some over-coverage at $T=2$. We might over-estimate the variance, but better than under-estimating it. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \cite{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov. rate"]), 3)}\%. Loess line is never flat.}

@@

@@

%% -- Optional special unnumbered sections -------------------------------------

>>>>

The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods. These tools will subsequently be implemented in an interactive evaluation framework for multiple imputation to aid applied researchers in drawing valid inference from incomplete datasets. This note provides the theoretical foundation towards the diagnostic evaluation of the convergence of MI algorithms. % for diagnostics for evalua one of these evaluation measures that is vital in evaluating MI procedures: a diagnostic to assess convergence of the MI algorithm. 
For reasons of brevity, this note only focuses on the MI algorithm implemented in \pkg{mice} \citep{mice}. %The implementation limits itself to the MI algorithm 
%More specifically, this note focuses on the MI algorithm that is implemented in the \proglang{R} package \pkg{mice}: `Multiple Imputation using Chained Equations' (MICE) \textbf{[can be removed, can be found in the mice ref]}. 
To this end, we will evaluate how convergence of the imputation algorithm can be diagnosed.

@@

The intended audience of this note consists of anyone who uses multiple imputation to solve missing data problems. Basic familiarity with MI methodology is assumed.\footnote{For the theoretical foundation of MI, see \cite{rubin87}. For an accessible and comprehensive introduction to MI from an applied perspective, see \cite{buur18}.}  
This note follows notation and conventions of \pkg{mice} \citep{mice}. % to suit the \pkg{mice} environment. %Deviations from the `original' notation by \cite{rubin87} are described in \citep[\S~2.2.3]{buur18}.

@@

To define $\widehat{R}$, we follow notation by \cite[p.~5]{veht19}. % The scalar summary of interest in this study is the average of the imputed values per variable. %As recommended by \cite{veht19} $\widehat{R}$ is computed as .... to be able to detect ... in the tails of the distribution. \cite[p.~5]{veth19}
Let $M$ be the total number of chains, $T$ the number of iterations per chain, and $\theta$ the scalar summary of interest (e.g., chain mean or chain variance). For each chain ($m = 1, 2, \dots, M$), we estimate the variance of $\theta$, and average these to obtain within-chain variance $W$.

@@

This study evaluates whether $\widehat{R}$ and $AC$ could diagnose convergence of multiple imputation algorithms. There is, however, no baseline measure available to evaluate their performance against.
\footnote{The current practice of visually inspecting imputation chains for signs of non-mixing or non-stationarity can only diagnose severely pathological cases of non-convergence.} 
We will assess the appropriateness of the two convergence diagnostics in comparison to evaluation criteria recommended by \cite[\S~2.5.2]{buur18}. The performance diagnostics comprise of average bias, average confidence interval width, and empirical coverage rate across simulations.% %"We expect that in accordance with Neyman (1934), at least 95 percent of the confidence intervals should contain the true population value, although some room for simulation error should be taken into account to counteract the finite nature of our simulations"\cite{schou18}.
\footnote{We could also look at distributional characteristics, and plausibility of imputed values, see \cite{vinknd} (n.d.). For now, this is outside of the scope of this study.} % Or use: Then evaluate the regular diagnostics of MI simulations, see Vink, n.d., and  convergence. Diagnostics include absolute bias of the estimated regression coefficient, confidence interval width, and empirical coverage rate across simulations. Convergence is evaluated with $\widehat{R}$, or potential scale reduction factor.

@@

\subsection{Data simulation}
The point of origin for all simulation conditions is a finite population of $N=1000$. The data are simulated to solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variable $X$, and covariates $Z_1$ and $Z_2$:

$$Y \sim \beta_1 X + \beta_2 Z_1 + \beta_3 Z_2.$$

The quantity of scientific interest is regression coefficient $\beta_1$. The data generating model is a multivariate normal distribution with means structure $\mu$ and variance-covariance matrix $\Sigma$.

@@

Outcome variable $Y$ is subsequently calculated as

$$Y =  2X + .5Z_1 - Z_2 + \epsilon .$$

@@

\subsection{Convergence Diagnostics}
% Add ``post-processing steps (e.g. aggregation computations, summary statistics, regressions on the simulation results) used to transform simulation outputs to reported results''. %, see \url{http://jsterman.scripts.mit.edu/docs/Rahmandad-Sterman%20Simulation%20Reporting%20Standards3.30%20update.pdf}.
We investigate convergence by extracting not only the imputed values of each MI procedure (the $T^{th}$ iteration), but also information ($\theta$-values) from the intermediate iterations ($t = 1, 2, \dots, T$). The $\theta$s of interest are chain means and chain variances per variable. We compute convergence diagnostics separately for each of the four variables in the dataset, and report the maximum (absolute) value as the `worst linear predictor' of convergence \textbf{[Rephrase or footnote??]}. $\widehat{R}$ is computed by implementing Vehtari et al.'s \citeyear{veht19} recommendations. $AC$ is computed with \pkg{stats} function \fct{acf}.

@@

We compute bias as the difference between $\bar{Q}$ and $Q$. Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval around $\bar{Q}$ (CI95\%). We compute the CI95\% bounds as

$$\bar{Q} \pm 2.66 \times SE_{\bar{Q}},$$

where 2.66 is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.

@@

\textbf{Empirical coverage rate seems more or less stable for conditions where $T>1$. We see some over-coverage at $T=2$. We might over-estimate the variance, but better than under-estimating it. On average, the coverage rate is somewhat higher than the expected nominal coverage of 95\% \citep{neym34}, namely \Sexpr{100*round(mean(dat[, "Cov. rate"]), 3)}\%. Loess line is never flat.}

@@

\textbf{Does this mean that conv is not nec to obtain correct inferences?}

@@

Also, conv diag for substantive model? wald test for AC = 0? 
%% -- Optional special unnumbered sections -------------------------------------
